# Loading R Packages {#loadingPackages}

There are three R packages available to Motus users:

1. motusClient: provides functions to download and update detections data and tag and receiver deployment metadata from the Motus server.

2. motus: provides functions to output summary plots, and to transform (add sun rise/sun set times), and analyze Motus data.

3. motusdata: provides sample datasets from the James Bay Shorebird Monitoring Project that we use throughout the book.

Motus **users** can install the latest stable versions of the R packages using the following code. You only need to install the packages once, but once installed, you do need to load each package (using library() or require()) each time you open a new R session. 

Please note that some functionalities of the devtools package may require updated versions of R and RStudio. To avoid errors, please ensure you are using the most recent releases of R (<https://www.r-project.org/>) and RStudio (<https://www.rstudio.com/products/RStudio/>), and update your R packages using update.pacakges() in the R console.


```{r update packages, eval = FALSE}
update.packages()                         ## to update your existing R packages
```

```{r install packages, eval = FALSE}
install.packages("devtools")              ## if you haven't already done this
library(devtools)                         ## load devtools

# install motusClient for data download
install_github("MotusWTS/motusClient@master")   ## the last stable version
library(motusClient)

# install motus for data manipulation, visualization and analysis
install_github("MotusWTS/motus@staging")   ## the last stable version
library(motus)

# install the motusdata package to get access to the sample data we use throughout this book
install_github("MotusWTS/motusdata")   ## the last stable version
library(motusdata)
```


Throughout the book, we use Tidyverse (<https://www.tidyverse.org/>), which is a collection of R packages for data science, including tidyr, dplyr, and ggplot2, and lubridate for managing and manipulating dates. These can be installed with:

```{r install tidyverse, eval = FALSE}
install.packages("tidyverse")             ## again, you only need to install once.
library(tidyverse)

install.packages("lubridate")            
library(lubridate)

```

More information on Tidyverse can be found at <https://www.tidyverse.org/>, or by browsing 'R for Data Science' by Garrett Grolemund and Hadley Wickham: <http://r4ds.had.co.nz/>.

## User Authentication {#userAuthentication}

The first time you call a function using the Motus R package, you will be asked for your motus.org username and password to authenticate your access to project data. This will only happen once per R session. If you do not have a Motus user name and password, you can sign up at <https://motus.org/data/user/new>. Permission to access project data will then be granted by Motus staff or the project principal investigator.

## Internal data processing {#internalProcessing}

As an animal moves within the detection range of a Motus station, radio transmissions, or 'bursts', are detected by antenna(s) and recorded by a receiver. These raw detection data are either uploaded to the Motus database instantaneously via internet connection, or downloaded from the receiver and uploaded to Motus manually. Behind the scenes, various functions read and process the raw detections data to produce the tag detections file that users access using the R package (see 'Data Import' below). While most users will not need to call on the internal data processing functions, a complete list of functions within the Motus server R package can be found on GitHub: <https://github.com/jbrzusto/motusServer>. The code behind each function can be viewed on GitHub, or by typing the following in the R console after loading the R package, replacing 'function.name' with the name of the R function of interest:

```{r function.name, eval = FALSE}

function.name() 

```

