# Loading Required Packages

Please note that some functionalities of the devtools package may require updated versions of R and RStudio. To avoid errors, please ensure you are using the most recent releases of R (<https://www.r-project.org/>) and RStudio (<https://www.rstudio.com/products/RStudio/>), and update your R packages using update.packages() in the R console.

```{r update packages, eval = FALSE}
update.packages()                         ## to update your existing R packages
```

```{r install packages, eval = FALSE}
install.packages("devtools")              ## if you haven't already done this
library(devtools)                         ## load devtools

```

Throughout the book, we use Tidyverse <https://www.tidyverse.org/>, which is a collection of R packages, including tidyr() and dplyr(). These can be installed with:

```{r install tidyverse, eval = FALSE}
install.packages("tidyverse")             ## again, you only need to install once.
library(tidyverse)
```

More information on Tidyverse can be found at <https://www.tidyverse.org/>, or by browsing 'R for Data Science' by Garrett Grolemund and Hadley Wickham: <http://r4ds.had.co.nz/>.


## Internal data processing

As an animal moves within the detection range of a Motus station, radio transmissions, or 'bursts', are detected by antenna(s) and recorded by a receiver. These raw detection data are either uploaded to the Motus database instantaneously via internet connection, or downloaded from the receiver and uploaded to Motus manually. Behind the scenes, various functions read and process the raw detections data to produce the tag detections file that users access using the R package (see 'Data Import' below). While most users will not need to call on the internal data processing functions, a complete list of functions within the Motus server R package can be found on GitHub: <https://github.com/jbrzusto/motusServer>. The code behind each function can be viewed on GitHub, or by typing the following in the R console after loading the R package, replacing 'function.name' with the name of the R function of interest:

```{r function.name, eval = FALSE}
function.name() 
```

