# Loading the Motus R Package

Motus **users** can install the latest stable version of the Motus R package using the following code. You only need to install the packages once, but once installed, you do need to load each package (using library() or require()) each time you open a new R session. 

Please note that some functionalities of the devtools package may require updated versions of R and RStudio. To avoid errors, please ensure you are using the most recent releases of R (<https://www.r-project.org/>) and RStudio (<https://www.rstudio.com/products/RStudio/>), and update your R packages using update.pacakges() in the R console.


```{r update packages, eval = FALSE}
update.packages()                         ## to update your existing R packages
```

```{r install packages, eval = FALSE}
install.packages("devtools")              ## if you haven't already done this
library(devtools)                         ## load devtools

install_github("jbrzusto/motus@master")   ## the last stable version
library(motus)
```


For **developers**, the 'staging' branch is for work-in-progress. Install it with:

```{r install packages developer, eval = FALSE}   
install.packages("devtools")               ## if you haven't already done this
library(devtools) 

install_github("jbrzusto/motus@staging")   ## the development version
library(motus)
```

Throughout the book, we use Tidyverse <https://www.tidyverse.org/>, which is a collection of R packages, including tidyr() and dplyr(). These can be installed with:

```{r install tidyverse, eval = FALSE}
install.packages("tidyverse")             ## again, you only need to install once.
library(tidyverse)
```

More information on Tidyverse can be found at <https://www.tidyverse.org/>, or by browsing 'R for Data Science' by Garrett Grolemund and Hadley Wickham: <http://r4ds.had.co.nz/>.

## User Authentication

The first time you call a function using the Motus R package, you will be asked for your motus.org username and password to authenticate your access to project data. This will only happen once per R session. If you do not have a Motus user name and password, you can sign up at <https://motus.org/data/user/new>. Permission to access project data will then be granted by Motus staff or the project principal investigator.

## Internal data processing

As an animal moves within the detection range of a Motus station, radio transmissions, or 'bursts', are detected by antenna(s) and recorded by a receiver. These raw detection data are either uploaded to the Motus database instantaneously via internet connection, or downloaded from the receiver and uploaded to Motus manually. Behind the scenes, various functions read and process the raw detections data to produce the tag detections file that users access using the R package (see 'Data Import' below). While most users will not need to call on the internal data processing functions, a complete list of functions within the Motus server R package can be found on GitHub: <https://github.com/jbrzusto/motusServer>. The code behind each function can be viewed on GitHub, or by typing the following in the R console after loading the R package, replacing 'function.name' with the name of the R function of interest:

```{r function.name, eval = FALSE}
function.name() 
```

