# Tag and Receiver Deployments {#deployments}

Before working with your detection data, a first step is to summarize and visualize the tag and receiver deployments registered to your project. Summarizing and plotting your deployments can be an effective way to find any errors in tag or receiver deployment metadata, which can in turn influence the completeness of the detections data for your project and the projects of others with detections of their own tags on your receivers. 

This chapter is a compliment to the online "notification center", which provides each project with a list of metadata issues (missing or outlying values) to be accepted or ignored. As such, please address any and all errors associated with your project in the notification center before importing your data through R. This chapter does not provide a full check of your deployment metadata, but will help uncover any errors that might be missed by the automatic queries in the notification center.

We use the James Bay Shorebird Project sample dataset throughout this chapter (see Section \@ref(sampleData); as you run through the code to look at your own deployments, **please fix any errors or omissions in your metadata by signing in to <https://motus.org/>**, and under the 'Manage Data' tab, select either 'Manage Tags' to fix tag deployment metadata or 'Manage Receivers' to fix receiver deployment metadata. It is important to fix metadata errors online, so that errors are fixed at the source and archived on the Motus Server, ensuring all users have access to the correct tag and receiver metadata.  

## Load relevant R packages and set working environment

```{r loadPackages.4, warning = FALSE, message = FALSE}

library(devtools)
library(tidyverse)
library(lubridate)    ## for working with dates
library(motusClient)
library(motus)    

## Set the system environment time zone to GMT (to ensure that you are always working in GMT)
Sys.setenv(TZ="GMT")

```

## Load .motus file

The .motus file has already been downloaded. To update and load into R, use tagme():

```{r loadDetections}

proj.num <- 176

sql.motus <- tagme(proj.num)

```

## Tag Deployments {#tagDeployments}

In your .motus file, you are provided with the metadata for tags registered to your own project, and metadata for duplicate tags from other projects which are associated with ambiguous detections in your data. We will check:

1. how many tags are registered to your project;
2. how many of those registered tags were deployed;
3. location of tag deployments;
4. completeness and accuracy of tag deployment metadata.

We will run through each of these in sequence.

### 1. Number of registered tags

To check the number of tags registered to your project, load the 'tags' table in the .motus file. The 'tags' table contains the metadata of each registered tag, including a unique tagID and information on manufacturer, model, nominal and offset frequency, burst interval, and pulse length. We select the metadata specific to the James Bay Shorebird Project, and ignore tag metadata associated with any duplicate tags belonging to other projects:


```{r importTags, message = FALSE, warning = FALSE}

tbl.tags <- tbl(sql.motus, "tags") 
df.tags <- tbl.tags %>%
                filter(projectID == proj.num) %>%
                collect() %>%
                as.data.frame()

```

The number of rows in the 'df.tags' database is equivalent to the number of tags registered to the James Bay Shorebird Project in the sample dataset:

```{r nRegisteredTags}

nrow(df.tags) ## number of registered tags in the database

```

### 2. Number of registered tags that were deployed

The tag deployment metadata table in the .motus file is required to check which registered tags have deployments. This file includes the date, time, species, and location of tag deployment. The database is subset to project '38', and we use the anti_join function to determine which registered tags have (or do not have) corresponding deployment information.

```{r importTagMeta.4, message = FALSE, warning = FALSE}

tbl.tagDeps <- tbl(sql.motus, "tagDeps") 
df.tagDeps <- tbl.tagDeps %>%
                filter(projectID == proj.num) %>%
                collect() %>%
                as.data.frame() %>% # once in dataframe format, can format dates using lubridate
                mutate(tsStart = as_datetime(tsStart, tz = "UTC", origin = "1970-01-01"),
                       tsEnd = as_datetime(tsEnd, tz = "UTC", origin = "1970-01-01")) 

anti_join(df.tags, df.tagDeps, by = "tagID") 

```

In the sample data, there are no registered tags without deployment metadata, which suggests that all tags were deployed. If you have undeployed tags in your own files, please check your records to ensure this is the case; without deployment metadata, detections for registered but 'undeployed' tags will be missing from your detections database.

### 3. Location of tag deployments

Creating a map of your tag deployments can point out any obvious errors in the tag deployment latitude or longitude that weren't captured by the online metadata message center queries.

a. **Install R mapping package.**

If you do not have rworldmap installed then you must install and load the package into R:

```{r installMapPackagesRecv1, message = FALSE, warning = FALSE}

install.packages("rworldmap")
install.packages("maproj")
require(rworldmap)
require(mapproj)

```

b. **Load base map files.**

```{r loadMapsRecv1, message = FALSE, warning = FALSE}

## get the lake data; adjust longitude
na.lakes <- map_data(map = "lakes") %>%
  mutate(long = long- 360)

## get the country data; adjust longitude
na.map <- filter(map_data(map="world2"),
                 region %in% c("Canada", "USA")) %>%
  mutate(long = long- 360)
                
## Others countries in the Americas that you may want to plot, depending on your location: "Mexico", "lakes","Belize", "Costa Rica", "Panama", "Guatemala", "Honduras", "Nicaragua", "El Salvador", "Colombia", "Venezuela", "Ecuador", "Peru", "Brazil", "Guyana","Suriname", "Bolivia", "French Guiana", "Jamaica", "Cuba", "Haiti", "Dominican Republic", "The Bahamas", "Turks and Caicos Islands", "Puerto Rico", "British Virgin Islands", "Montserrat", "Dominica", "Saint Lucia", "Barbados", "Grenada", "Trinidad and Tobago", "Chile", "Argentina", "Uruguay"

```


c. **Map the locations of tag deployments.**

Map the location of tag deployments for the sample data: 

```{r mapRecvs1}

## set limits to map based on locations of detections, ensuring they include the deployment locations
xmin <- -100 #min(df.tagDeps$longitude, na.rm = TRUE) - 5
xmax <- max(df.tagDeps$longitude, na.rm = TRUE) + 5
ymin <- min(df.tagDeps$latitude, na.rm = TRUE) - 5
ymax <- max(df.tagDeps$latitude, na.rm = TRUE) + 5
                
## map using ggplot
ggplot(na.lakes, aes(long, lat)) + 
  geom_polygon(
    data = na.map, 
    aes(long, lat, group=group), colour = "grey", fill="grey98") + 
  geom_polygon(
    aes(group = group), colour = "grey", fill = "white") +
  coord_map(projection="mercator", 
            xlim = c(xmin, xmax), 
            ylim = c(ymin, ymax)) +
  xlab("") + ylab("") + 
  theme_bw() + 
  geom_point(data = df.tagDeps, 
             aes(longitude, latitude), cex = 2, pch = 1, colour = "red")
   
```

### 4. Check completeness and accuracy of tag deployment metadata

Required tag metadata includes deployment start date/time, end date/time (if applicable), deployment latitude, deployment longitude, and species. Lack of information on deployment date, time, and location in particular can influence the estimated lifespan of your tag, and therefore whether the tagFinder will 'look' for your tag at the appropriate time(s). It can also increase the potential for ambiguities with duplicate tags in the system. 

a. **Look at range of metadata values**.

As a first step, use summary(df.tagDeps) to get an idea of the range of each variable, and whether any variables have missing (NA) or odd values in the data. The following summarizes a subset of the variables in the df.tagDeps database. There are several things to consider: are the range of start and end dates reasonable for your deployments, or are there obvious errors in the timing of deployments? Is the range in deployment latitude and longitude reasonable? Are the values for species IDs correct?  

```{r summaryTagMeta}

df.tagDeps %>%
      select(tagID, projectID, tsStart, tsEnd, speciesID, latitude, longitude) %>%
      summary()

```

There are no missing start and end dates (tsStart/tsEnd), and deployment start dates range from `r min(year(df.tagDeps$tsStart))` to `r max(year(df.tagDeps$tsStart))`, which is reasonable for this project.  

The species IDs are numeric, and somewhat meaningless without an ability to assign an actual species name to the numeric ID, which we do next. 

b. **Check that species IDs are appropriate for your data**.

The 'species' table in the .motus file associates each numeric species ID with an English, French, and scientific name. We load that table, and subset to the suite of numeric speciesIDs in your tag metadata:

```{r checkSpecies}

## list of species IDs in project 38 metadata
sp.list <- unique(df.tagDeps$speciesID)     ## generate list of speciesIDs in the tag metadata

## Species metadata
tbl.species <- tbl(sql.motus, "species") 
tbl.species %>%
    filter(id %in% sp.list) %>%
    collect() %>%
    as.data.frame()

```

This lists all species that are included in the tag deployment metadata for the project. If there are species that do not make sense, this is likely due to a data entry error when assigning a deployment to a species. You can look for records in your tag metadata that are associated with a particular speciesID using the following code; you would then use the deployID associated with the entry/entries to find and update the deployment record in your project metadata online:

```{r listMetaSpecies}

filter(df.tagDeps, speciesID == 4780)

```

## Check Receiver Metadata {#recvMetadata}

There are two sources of receiver metadata in Motus detection data: receivers registered to your own project, and receivers registered to the projects of others. You are provided access to the metadata for all receivers in the network, because negative data (i.e., my tag was *not* detected at tower x even though tower x was active) is often as important as positive data. It also allows you to map where your tags were detected relative to the distribution of receivers throughout the Motus network.

Receiver metadata errors or omissions that you find in your .motus file can only be fixed for receivers registered to your own project. 

All users are encouraged to enter complete and accurate receiver metadata for the benefit of the entire network. If you anticipate needing specific information on receiver or antenna deployments for towers deployed by others, please consider using the Motus discussion group (<https://motus.org/discussion/>) to request that other registered users record the receiver deployment details you will need; be specific about the exact receiver deployment details you are interested in, and when and where in the network your tags will be deployed and potentially detected. 

In the following steps we will check:

1. number of project receiver deployments;
2. timing of project receiver deployments;
3. location of network-wide and project receiver deployments;
4. completeness and accuracy of receiver metadata.

### 1. Number of project receiver deployments

To see which (if any) receiver deployments are registered to your project, import, subset and summarize the receiver deployment data:

```{r projectDeps}

tbl.recvDeps <- tbl(sql.motus, "recvDeps") 
df.projRecvs <- tbl.recvDeps %>%
                filter(projectID == proj.num) %>%
                collect() %>%
                as.data.frame() %>%
                mutate(tsStart = as_datetime(tsStart, tz = "UTC", origin = "1970-01-01"),
                       tsEnd = as_datetime(tsEnd, tz = "UTC", origin = "1970-01-01"))

summary(df.projRecvs)

```

There are `r nrow(df.projRecvs)` receiver deployments registered to the sample project. Three deployments are missing latitude and longitude, and two deployments are missing end dates, which suggests that those receivers are still deployed. 

The following code keeps only parameters of interest, and arranges the remaining records by receiver ID, latitude, and start date:

```{r checkRegisteredReceivers}

df.projRecvs %>%
  arrange(deviceID, latitude, tsStart) %>%
  mutate(dateStart = date(tsStart)) %>%
  select(-serno,-fixtureType, -macAddress, isRecvMobile, -tsStart, -tsEnd, -elevation, - projectID, -status, -receiverType) 

```

The number of receiver deployments in the metadata should correspond with the number of field deployments. 

Looking at the 'isRecvMobile' column for the three receiver deployments that are missing latitude and longitude information, it is evident that these are mobile receivers which do not have a fixed position, and so it is not possible to assign a latitude and longitude (these fields will remain NA). 

### 2. Timing of project receiver deployments

The timing of deployments can be displayed graphically; horizontal line(s) in the following plot show the time span for each receiver (deviceID) deployment registered to the James Bay Shorebird Project. Note that for the two receivers without deployment end dates, the code assigns an arbitrary end date based on the maximum end date of the other receivers plus one month - without this fix, deployments without end dates do not get displayed. Different deployments of the same receiver should not overlap in time:

```{r projectRecvDeploy, warnings = FALSE, messages = FALSE}

# put data in long format first

df.projRecvs.long <- select(df.projRecvs, deviceID, deployID, tsStart, tsEnd) %>% gather(when, ts,c(tsStart, tsEnd))

ggplot(df.projRecvs.long, aes(y = as.factor(deviceID), x = if_else(is.na(ts), (max(ts, na.rm = TRUE) + 2678400), ts ), colour = as.factor(deployID))) +
  geom_line() + theme_bw() +
  ylab("Receiver ID") + xlab("Year") + scale_colour_discrete(guide = guide_legend(title ="Receiver deployment ID"))

```

### 3. Location of receiver deployments

Maps provide better spatial context than simple plots; the following steps plot the location of Motus receivers on a map of North America, with receivers deployed by the sample project displayed in red.

a. **Install R mapping package.**

Install and load the 'rworldmap' package into R:

```{r installMapPackagesRecv2, message = FALSE, warning = FALSE}

install.packages("rworldmap")
require(rworldmap)

```

b. **Load base map files.**

```{r loadMapsRecv2, message = FALSE, warning = FALSE}

na.lakes <- map_data(map = "lakes")
na.lakes <- mutate(na.lakes, long = long- 360)

# Include all of the Americas to begin

na.map <- subset(map_data(map="world2"), 
                region %in% c("Canada", "USA", "Mexico", "lakes",
                               "Belize", "Costa Rica", "Panama", 
                               "Guatemala", "Honduras", "Nicaragua", 
                               "El Salvador", "Colombia", "Venezuela", "Ecuador", "Peru", "Brazil",
                               "Guyana","Suriname", "Bolivia", "French Guiana", "Jamaica", "Cuba", 
                               "Haiti", "Dominican Republic", "The Bahamas", "Turks and Caicos Islands", 
                               "Puerto Rico", "British Virgin Islands", "Montserrat", "Dominica", "Saint Lucia", 
                               "Barbados", "Grenada", "Trinidad and Tobago", "Chile", "Argentina", 
                               "Uruguay"))

na.map <- mutate(na.map, long = long- 360)

```

c. **Map the location of receivers in the Americas.**

Map showing the location of network-wide receivers (dark grey 'x') and receivers deployed by the James Bay Shorebird Project (project 38; red 'x').

```{r mapRecvs2, message = FALSE, warning = FALSE}

# set limits to map based on locations of detections, ensuring they include the deployment locations
xmin <- min(df.recvDeps$longitude, na.rm = TRUE) - 2
xmax <- -20 # restrict to the Americas (excluding a few points in Europe)
ymin <- min(df.recvDeps$latitude, na.rm = TRUE) - 2
ymax <- max(df.recvDeps$latitude, na.rm = TRUE) + 2
                
# map
ggplot(na.lakes, aes(long, lat))+ 
  geom_polygon(data = na.map, aes(long, lat, group=group), colour = "grey", fill="grey98")+#
  geom_polygon(aes(group = group), colour = "grey", fill = "white")+
  coord_map(projection="mercator", xlim = c(xmin, xmax), ylim = c(ymin, ymax))+
  xlab("") + ylab("") + 
  theme_bw() + 
  geom_point(data = df.recvDeps, aes(longitude, latitude, colour = as.logical(projectID == 38)), cex = 0.8, pch = 4)+
  scale_colour_manual(values = c("grey30", "red"), name = "Project 38 Deployment") 
  

```

d. **Map the location of project specific receivers only.**

Map of project-specific receivers, created by setting the x-axis (longitude) and y-axis (latitude) map limits using the 'df.projRecvs' dataframe created above. Deployments are restricted to those that were active at in 2016.

```{r mapProjRecvs, message = FALSE, warning = FALSE}

# set limits to map based on locations of detections, ensuring they include the deployment locations
xmin <- min(df.projRecvs$longitude, na.rm = TRUE) - 2
xmax <- max(df.projRecvs$longitude, na.rm = TRUE) + 2
ymin <- min(df.projRecvs$latitude, na.rm = TRUE) - 1
ymax <- max(df.projRecvs$latitude, na.rm = TRUE) + 1
                
# map
ggplot(na.lakes, aes(long, lat))+ 
  geom_polygon(data = na.map, aes(long, lat, group=group), colour = "grey", fill="grey98")+#
  geom_polygon(aes(group = group), colour = "grey", fill = "white")+
  coord_map(projection="mercator", xlim = c(xmin, xmax), ylim = c(ymin, ymax))+
  xlab("") + ylab("") + 
  theme_bw() + 
  geom_point(data = filter(df.projRecvs, year(tsStart) >= 2016 & year(tsEnd) <= 2016), aes(longitude, latitude, colour = as.factor(deviceID)), cex = 2, pch = 1)+
  scale_colour_discrete(name  =  "Receiver ID") 

```

### 4. Completeness and accuracy of receiver metadata

Motus users will be concerned primarily with the completeness of metadata for receiver deployments with detection(s) of their tags, because these can directly influence the interpretation of those detections. For example, missing deployment latitude or longitude will result in an unknown location for the tag detection, and missing information on antenna type and/or orientation can impede the estimation of flight or departure orientation. 

In many cases, however, metadata for receiver deployments *without* tag detections can also be useful, for example to estimate probability of detecting an animal that passes within range of a station.

In this section, the focus is on metadata for receivers registered to a particular project. Depending on your interests, these summaries can be applied to a larger group of receivers, e.g., all receivers with detections or all receivers within certain geographic limits (with or without detections).

a. **Load receiver and antenna metadata**

```{r loadReceiverAntennaMetadata}

# antenna metadata for ALL Motus antenna deployments; keep only variables we are interested in to simplify
tbl.antDeps <- tbl(t, "antDeps") 
df.antDeps <- tbl.antDeps %>%
                select(deployID, port, antennaType, bearing, heightMeters) %>%
                collect() %>%
                as.data.frame()

# receiver deployments; select parameters of interest
df.recvDeps <- df.recvDeps %>%
                    select(deployID, receiverType, deviceID, name, latitude, longitude, isRecvMobile, tsStart, tsEnd, projectID, elevation) 

df.stationDeps <- left_join(df.recvDeps, df.antDeps, by = "deployID")

```

Subset these to receivers registered to a project:

```{r stationMetaProj}

df.stationDeps <- filter(df.stationDeps, projectID == proj.num)

```

b. **Look at range of metadata values**. 

Use summary() to get a general idea of the distribution of the variables in the data. 

```{r SummaryRecv}

summary(df.stationDeps)

```

There are the 4 deployments with missing latitude and longitude associated with the four deployments of mobile receivers that we saw earlier. 

Elevation is missing from 53 of 60 records, but elevation is not a required field, and can be estimated from other sources, or directly in R (for example, see <https://stackoverflow.com/questions/8973695/conversion-for-latitude-longitude-to-altitude-in-r>). 

Antenna bearing is missing from 13 of 60 records, and height of the antenna(s) is missing for 4 of 60 records.  Subset the records with missing antenna bearing to see if these can be fixed:

```{r antennaBearing}

filter(df.stationDeps, is.na(bearing)) %>%
  select(-elevation, -deviceID, -tsEnd)

```

Receiver deployments with missing antenna bearing(s) are restricted to deployments of omni-directional antennas or mobile receivers, and so the missing values make sense. These records also show that the four records with missing antenna height are also associated with the four mobile receivers, and so again the missing values make sense and cannot be fixed.

