# Tag and Receiver Deployments {#deployments}

Before working with your detections data, a first step is to summarize and visualize the tag and receiver deployments registered to your project. Summarizing and plotting your deployments can be an effective way to find any errors in tag or receiver deployment metadata, which can in turn influence the completeness of the detections data for your project and the projects of others with detections of their own tags on your receivers. 

This chapter is a compliment to the online "message center", which provides each project with a list of metadata issues (missing or outlying values) to be accepted or ignored. As such, please address any and all errors associated with your project in the message center before importing your data with R. This chapter does not provide a full check of your deployment metadata, but will help uncover any errors that might be missed by the automatic queries in the message center.

We use the James Bay Shorebird Project sample dataset throughout this chapter (see Section \@ref(sampleData); as you run through the code to look at your own deployments, **please fix any errors or ommissions in your metadata by signing in to <https://motus.org/>**, and under the 'Manage Data' tab, select either 'Manage Tags' to fix tag deployment metadata or 'Manage Receivers' to fix receiver deployment metadata. It is important to fix metadata errors online, so that errors are fixed at the source and archived on the Motus Server, ensuring all users have access to the correct tag and receiver metadata.  

## Load relevant R packages

```{r loadPackages.4, warning = FALSE, message = FALSE}

library(devtools)
library(tidyverse)
library(lubridate)    ## for working with dates
library(motusdata)    ## repository for sample datasets for this book

```

## Tag Deployments {#tagDeployments}

In your .motus file, you are provided with the metadata for tags registered to your own project, and for duplicate tags from other projects associated with ambiguous detections in your data. We will check:

1. how many tags are registered to your project;
2. how many of those registered tags were deployed;
3. location of tag deployments;
4. timing of tag deployments;
3. completeness and accuracy of tag deployment metadata.

We will run through each of these in sequence.

### Number of registered tags

To check the number of tags registered to your project, load the 'df.tags' RDS file for the James Bay Shorebird Monitoring Project, which was created using the 'createFlatFiles' function in Section \@ref(flatRDSfunction). The file is included for your use in the motusdata package. 

The 'df.tags' file contains the characteristics of each registered tag, including a unique tagID and information on manufacturer, model, nominal and offset frequency, burst interval, and pulse length. We select the metadata specific to the James Bay Shorebird Monitoring Project, and ignore tag metadata associated with any duplicate tags belonging to other projects:


```{r importTags, message = FALSE, warning = FALSE}

df.tags <- readRDS("./data/df.tags.RDS")
df.tags <- df.tags %>% filter(projectID == 38)

```

The number of rows in the 'df.tags' database is equivalent to the number of tags registered on Motus to the James Bay Shorebird Project. There are 691 tags registered to the project (recall we are using just a subset of this dataset):

```{r nRegisteredTags}

nrow(df.tags) ## number of registered tags in the database

```

### Number of registered tags that were deployed

The tag deployment table ('df.tagDeps') is required to check which registered tags have deployments. This file includes the date, time, species, and location of tag deployment. The database is subset to project '38', and selected columns are merged with the 'df.tags' table to determine which registered tags have (or do not have) corresponding deployment information.

```{r importTagMeta.4, message = FALSE, warning = FALSE}

df.tagDeps <- readRDS("./data/df.tagDeps.RDS")                                        
df.tagDeps <- filter(df.tagDeps, projectID == 38) 

## merge with tag registration metadata:

df.tagsDeployed <- full_join(df.tags, select(df.tagDeps, deployID, tagID), by = "tagID")

```

To list registered tags that do not have associated deployment metadata:

```{r tagsRegNotDeployed}

filter(df.tagsDeployed, is.na(deployID))

```

Two of the 691 registered tags do not have deployment metadata, which suggests that they were not deployed. Records should be checked to ensure this is the case; without deployment metadata, detections for registered but 'undeployed' tags will be missing from your detections database.

### Location of tag deployments

Creating a map of your tag deployments can point out any obvious errors in the tag deployment latitude or longitude that aren't captured by the online metadata message center queries.

1. **Install R mapping package.**

If you do not have rworldmap installed then you must install and load the package into R first:

```{r installMapPackagesRecv, message = FALSE, warning = FALSE}

install.packages("rworldmap")
require(rworldmap)

```

2. **Load base map files.**

```{r loadMapsRecv, message = FALSE, warning = FALSE}

na.lakes <- map_data(map = "lakes")
na.lakes <- mutate(na.lakes, long = long- 360)

## Include the countries that make sense

na.map <- subset(map_data(map="world2"), 
                region %in% c("Canada", "USA")) 
                
## Others (Americas): "Mexico", "lakes","Belize", "Costa Rica", "Panama", "Guatemala", "Honduras", "Nicaragua", "El Salvador", "Colombia", "Venezuela", "Ecuador", "Peru", "Brazil", "Guyana","Suriname", "Bolivia", "French Guiana", "Jamaica", "Cuba", "Haiti", "Dominican Republic", "The Bahamas", "Turks and Caicos Islands", "Puerto Rico", "British Virgin Islands", "Montserrat", "Dominica", "Saint Lucia", "Barbados", "Grenada", "Trinidad and Tobago", "Chile", "Argentina", "Uruguay"

na.map <- mutate(na.map, long = long- 360)

```

3. **Map the locations of tag deployments.**

Map the location of tag deployments for project 38: 

```{r mapRecvs}

## set limits to map based on locations of detections, ensuring they include the deployment locations
xmin <- -100 #min(df.tagDeps$longitude, na.rm = TRUE) - 5
xmax <- max(df.tagDeps$longitude, na.rm = TRUE) + 5
ymin <- min(df.tagDeps$latitude, na.rm = TRUE) - 5
ymax <- max(df.tagDeps$latitude, na.rm = TRUE) + 5
                
## map
ggplot(na.lakes, aes(long, lat))+ 
  geom_polygon(data = na.map, aes(long, lat, group=group), colour = "grey", fill="grey98")+#
  geom_polygon(aes(group = group), colour = "grey", fill = "white")+
  coord_map(projection="mercator", xlim = c(xmin, xmax), ylim = c(ymin, ymax))+
  xlab("") + ylab("") + 
  theme_bw() + 
  geom_point(data = df.tagDeps, aes(longitude, latitude), cex = 2, pch = 1, colour = "red")
   
```
There is a warning that 164 rows were removed from the map due to missing [latitude and/or longitude] values. Summarizing the data with `r summary(df.tagDeps), eval = FALSE` suggests that 164 deployments are missing speciesID, latitude, and longitude. Comparing the tagIDs for these records in the 'df.tagDeps' file with the deployments of those tags online at <https://motus.org>, these tags were registered but were either not deployed or were used as test tags. 

### Check accuracy of tag deployment metadata

Required tag metadata includes deployment start date/time, end date/time (if applicable), deployment latitude, deployment longitude, and species. Lack of information on deployment date, time, and location in particular can influence the estimated lifespan of your tag, and therefore whether the tagFinder will 'look' for your tag at the appropriate time(s). It can also increase the potential for ambiguities with duplicate tags in the system. 

1. **Look at range of metadata values**.

As a first step, use summary(df.tagDeps) to get an idea of the range of each variable, and whether any variables have missing (NA) or odd values in the data. The following summarizes a subset of the variables in the df.tagDeps database. There are several things to consider: are the range of start and end dates reasonable for your deployments, or are there obvious errors in the timing of deployments? Is the range in deployment latitude and longitude values reasonable? Are the values for species IDs correct?  

```{r summaryTagMeta}

summary(select(df.tagDeps, deployID, tagID, projectID, tsStart, tsEnd, speciesID, latitude, longitude))

```

There are no missing start and end dates (tsStart/tsEnd), and deployment start dates range from `r min(year(df.tagDeps$tsStart)` to `r max(year(df.tagDeps$tsStart)`, which is reasonable for this project. However, as we saw above, there are `r length(df.tagDeps$speciesID[is.na(df.tagDeps$speciesID)])` records missing speciesID, latitude, and longitude. 

The species IDs are numeric, and somewhat meaningless without an ability to assign an actual species name to the numeric ID, which we do next. 

2. **Check that species IDs are appropriate for your data**.

The 'df.species' RMD file associates each numeric species ID with an english, french, and scientific name. We import that table, and subset to the suite of numeric speciesIDs in your tag metadata:

```{r checkSpecies}

## Species metadata
df.species <- readRDS("./data/df.species.rds")

## list of species IDs in project 38 metadata
sp.list <- unique(df.tagDeps$speciesID)     ## generate list of speciesIDs in the tag metadata
df.species %>% 
        filter(id %in% sp.list) %>% 
        collect() %>% 
        as.data.frame()                     ## subset and convert to a flat file

```

This lists all species that are included in the tag deployment metadata for the project. If there are species that do not make sense, this is likely due to a data entry error when assigning a deployment to a species. You can look for records in your tag metadata that are associated with a particular speciesID using the following code; you would then use the deployID associated with the entry/entries to find and update the deployment record in your project metadata online:

```{r listMetaSpecies}

filter(df.tagDeps, speciesID == 5000)

```

## Check Receiver Metadata {#recvMetadata}

There are two sources of receiver metadata in Motus detections data: receivers registered to your own project, and receivers registered to the projects of others. You are provided access to the metadata for all receivers in the network, because negative data (i.e., my tag was *not* detected at tower x even though it was active) is often as important as positive data. It also allows you to map where your tags were detected relative to the distribution of receivers throughout the Motus network.

Receiver metadata errors or ommissions that you find in your .motus file can only be fixed for receivers registered to your own project. All users are encouraged to enter complete and accurate receiver metadata for the benefit of the entire network. If you anticipate needing specific information on receiver or antenna deployments for your project, please consider using the Motus listserve to request that other registered users record the receiver deployment details you will need; be specific about the exact receiver deployment details you are interested in, and when and where in the network your tags will be deployed and potentially detected. 

In the following steps will check:

1. number of project receiver deployments;
2. timinig of project receiver deployments;
3. location of network-wide and project receiver deployments;
4. completeness and accuracy of receiver metadata.

### Number of project receiver deployments

To see which (in any) receiver deployments are registered to your project, import, subset and summarize the deployment data:

```{r projectDeps}

df.recvDeps <- readRDS("./data/df.recvDeps.rds")
df.projRecvs <- filter(df.recvDeps, projectID == 38)     # replace with your own projectID for your own data. 
summary(df.projRecvs)

```

There are `r nrow(df.projRecvs)` receiver deployments registered to project 38. Four deployments are missing latitude and longitude, and two deployments are missing end dates, which suggests those receivers are still deployed. 

The following code keeps only parameters of interest, and arranges the remaining records by receiver ID, latitude, and start date:

```{r checkRegisteredReceivers}

df.projRecvs %>%
  arrange(deviceID, latitude, tsStart) %>%
  mutate(dateStart = date(tsStart)) %>%
  select(-serno,-fixtureType, -macAddress, isMobile, -tsStart, -tsEnd, -elevation, - projectID, -status, -receiverType) 

```

The number of receiver deployments in the metadata should correspond with the number of field deployments. Looking at the 'name' and 'isMobile' columns for the four receiver deployments that are missing latitude and longitude information, it is evident that these are mobile receivers which do not have a fixed position. The parameter'isMobile' = 0 for deployIDs 3813 and 3814, despite being named as mobile receivers; the classification of these receiver deployments should be changed to 'isMobile' = 1 online at <https://motus.org> under receiver deployments for this project.

### Timing of project receiver deployments

The timing of deployments can be displayed graphically; horizontal line(s) in the following plot show the time span for each deployment for each receiver (deviceID) registered to the James Bay Shorebird Monitoring project. Different deployments of the same receiver should not overlap in time:

```{r projectRecvDeploy, warnings = FALSE, messages = FALSE}

# put data in long format first

df.projRecvs.long <- select(df.projRecvs, deviceID, deployID, tsStart, tsEnd) %>% gather(when, ts,c(tsStart, tsEnd))

ggplot(df.projRecvs.long, aes(y = as.factor(deviceID), x = ts, colour = as.factor(deployID))) +
  geom_line() + theme_bw() +
  ylab("Receiver ID") + xlab("Year") + scale_colour_discrete(guide = guide_legend(title ="Receiver deployment ID"))

```

### Location of receiver deployments

#### Map of receiver locations

Maps provide better spatial context than simple plots; the following steps plot the location of Motus receivers on a map of North America, with receivers deployed by project 38 displayed in red.

1. **Install R mapping package.**

Install and load the 'rworldmap' package into R:

```{r installMapPackagesRecv, message = FALSE, warning = FALSE}

install.packages("rworldmap")
require(rworldmap)

```

2. **Load base map files.**

```{r loadMapsRecv, message = FALSE, warning = FALSE}

na.lakes <- map_data(map = "lakes")
na.lakes <- mutate(na.lakes, long = long- 360)

# Include all of the Americas to begin

na.map <- subset(map_data(map="world2"), 
                region %in% c("Canada", "USA", "Mexico", "lakes",
                               "Belize", "Costa Rica", "Panama", 
                               "Guatemala", "Honduras", "Nicaragua", 
                               "El Salvador", "Colombia", "Venezuela", "Ecuador", "Peru", "Brazil",
                               "Guyana","Suriname", "Bolivia", "French Guiana", "Jamaica", "Cuba", 
                               "Haiti", "Dominican Republic", "The Bahamas", "Turks and Caicos Islands", 
                               "Puerto Rico", "British Virgin Islands", "Montserrat", "Dominica", "Saint Lucia", 
                               "Barbados", "Grenada", "Trinidad and Tobago", "Chile", "Argentina", 
                               "Uruguay"))

na.map <- mutate(na.map, long = long- 360)

```

3. **Map the locations of receivers in the Americas.**

Map showing the location of network-wide receivers (dark grey 'x') and receivers deployed by the James Bay Shorebird Project (project 38; red 'x').

```{r mapRecvs, message = FALSE, warning = FALSE}

# set limits to map based on locations of detections, ensuring they include the deployment locations
xmin <- min(df.recvDeps$longitude, na.rm = TRUE) - 2
xmax <- -20 # restrict to the Americas (excluding a few points in Europe)
ymin <- min(df.recvDeps$latitude, na.rm = TRUE) - 2
ymax <- max(df.recvDeps$latitude, na.rm = TRUE) + 2
                
# map
ggplot(na.lakes, aes(long, lat))+ 
  geom_polygon(data = na.map, aes(long, lat, group=group), colour = "grey", fill="grey98")+#
  geom_polygon(aes(group = group), colour = "grey", fill = "white")+
  coord_map(projection="mercator", xlim = c(xmin, xmax), ylim = c(ymin, ymax))+
  xlab("") + ylab("") + 
  theme_bw() + 
  geom_point(data = df.recvDeps, aes(longitude, latitude, colour = as.logical(projectID == 38)), cex = 0.8, pch = 4)+
  scale_colour_manual(values = c("grey30", "red"), name = "Project 38 Deployment") 
  

```

4. **Map the locations of project specific receivers only.**

Map of project-specific receivers, created by setting the x-axis (longitude) and y-axis (latitude) map limits using the 'df.projRecvs' dataframe created above. Deployments are restricted to those that were active at in 2016.

```{r mapProjRecvs, message = FALSE, warning = FALSE}

# set limits to map based on locations of detections, ensuring they include the deployment locations
xmin <- min(df.projRecvs$longitude, na.rm = TRUE) - 2
xmax <- max(df.projRecvs$longitude, na.rm = TRUE) + 2
ymin <- min(df.projRecvs$latitude, na.rm = TRUE) - 1
ymax <- max(df.projRecvs$latitude, na.rm = TRUE) + 1
                
# map
ggplot(na.lakes, aes(long, lat))+ 
  geom_polygon(data = na.map, aes(long, lat, group=group), colour = "grey", fill="grey98")+#
  geom_polygon(aes(group = group), colour = "grey", fill = "white")+
  coord_map(projection="mercator", xlim = c(xmin, xmax), ylim = c(ymin, ymax))+
  xlab("") + ylab("") + 
  theme_bw() + 
  geom_point(data = filter(df.projRecvs, year(tsStart) >= 2016 & year(tsEnd) <= 2016), aes(longitude, latitude, colour = as.factor(deviceID)), cex = 2, pch = 1)+
  scale_colour_discrete(name  =  "Receiver ID") 

```

### Completeness and accuracy of receiver metadata

Motus users will be concerned primarily with the completeness of metadata for receiver deployments with detection(s) of their tags, because these can directly influence the interpretation of those detections. For example, missing deployment latitude or longitude will result in an unknown location for the tag detection, and missing information on antenna type and/or orientation can impede the estimation of flight or departure orientation. 

In many cases, however, metadata for receiver deployments *without* tag detections can also be useful, for example to estimate probability of detecting an animal that passes within range of a station.

In this section, the focus is on metadata for receivers registered to a particular project, in this case project 38. Depending on your interests, these summaries can be applied to a larger group of receivers, e.g., all receivers with detections or all receivers within certain geographic limits (with or without detections).

1. **Load receiver and antenna metadata**

```{r loadReceiverAntennaMetadata}

# antenna metadata for ALL Motus antenna deployments
df.antDeps <- readRDS("./data/df.antDeps.rds")

# keep only variables that we are currently interested in, to simplify:
df.antDeps <- df.antDeps %>%
                mutate(ant = port) %>%
                select(deployID, ant, antennaType, bearing, heightMeters)

df.recvDeps <- df.recvDeps %>%
                    select(deployID, receiverType, deviceID, name, latitude, longitude, isMobile, tsStart, tsEnd, projectID, elevation) 

df.stationDeps <- left_join(df.recvDeps, df.antDeps, by = "deployID")

```

Subset these to receivers registered to a project, in this case project 38:

```{r stationMetaProj}

df.stationDeps <- filter(df.stationDeps, projectID == 38)

```

2. **Look at range of metadata values**. 

As with the tag metadata, use summary() to get a general idea of the distribution of the variables in the data. 

```{r SummaryRecv}

summary(df.stationDeps)

```

There are the 4 deployments with missing latitude and longitude associated with the four deployments of mobile receivers that we saw earlier. Elevation is missing from 53 of 60 records, but elevation is not a required field, and can be estimated from other sources, or directly in R (for example, see <https://stackoverflow.com/questions/8973695/conversion-for-latitude-longitude-to-altitude-in-r>). Antenna bearing is missing from 13 of 60 records, and height of the antenna(s) is missing for 4 of 60 records.  Subset the records with missing antenna bearing to see if these can be fixed:

```{r antennaBearing}

filter(df.stationDeps, is.na(bearing)) %>%
  select(-elevation, -deviceID, -tsEnd)

```

Receiver deployments with missing antenna bearing(s) are restricted to deployments of omni-directional antennas or mobile receivers, and so the missing values make sense. These records also show that the four records with missing antenna height are also associated with the four mobile receivers, and so again the missing values make sense and can't be fixed.

